---
output: html_document
chapter: Calculating Slopes (in progress)
editor_options:
  chunk_output_type: console
---

# Calculating Linear Slopes

In this chapter, we will use a fake data set that I have generated by eyeballing Figure 3 from this paper ([PDF](https://www.smin95.com/pubs/mao2020.pdf)):

**Yu Mao, Seung Hyun Min, Shijia Chen, Ling Gong, Hao Chen, Robert F. Hess, Jiawei Zhou. Binocular imbalance in amblyopia depends on spatial frequency in binocular combination. IOVS. 2020;61(8):7.**

This data set contains both subject groups (`Normal` and `Amblyopia`) and conditions (`Condition`). 

Let's begin by loading the `tidyverse` and other libraries, and uploading the csv file `amblyopia_random.csv` using `read_csv()` from the `tidyverse` package.

It is always a good habit to make sure that the data set you intended to load uploads properly by using `head()`, which returns the first 6 rows of the data frame. 

```{r}
library(tidyverse)
library(smplot)
library(cowplot)
df <- read_csv('https://www.smin95.com/amblyopia_random.csv') 
head(df)
```

In this dataset, there are five columns.

- First, `Subject` column has all the subjects. People with amblyopia (a visual disorder) are labelled with `A`. First subject with amblyopia is written as `A1`. Normal subject is written with `N`; first normal subject is `N1`. We see that there are 10 subjects per group, so there are 20 subjects total.

```{r}
unique(df$Subject)
```

- Second, `absBP` is the data of our interest. 

- Third, `SF` refers to spatial frequency. We will be calculating slopes of `absBP` as a function of `SF`. So these are our x-coordinates for slope calculations (see sections below). Each unit increases by a factor of 2, so it is also helpful to convert the values into log2 units. 

```{r}
length(unique(df$SF))
df <- df %>% mutate(logSF = log2(SF))
head(df)
```

- Fourth, `Group` refers to as the subject gruop. There are two groups: `Amblyopia` and `Normal`.

```{r}
unique(df$Group)
```

- Lastly, `Condition` refers to the testing condition. In this dataset, there are two conditions.

```{r}
unique(df$Condition)
```

The columns `Group` and `Condition` are categorical variable and must therefore be factors. `head(df)` shows that `Group` and `Condition` are `<chr>`, which mean characters. Lets change them to factors `<fct>`. 

```{r}
df$Group <- factor(df$Group)
df$Condition <- factor(df$Condition)
head(df)
```

We see that `Group` and `Condition` columns are now factor `<fct>`. 

## Linear relationship using `lm()`

Linear relationship between $x$ and $y$ can be described as $y = mx + b$, where y is the dependent variable, x is the independent variable, m is the slope, and b is the y-intercept.

Let's try calculate $m$ and $b$ using the data of `A3`, which is the 3rd amblyopic observer.

```{r}
A3 <- df %>% filter(Subject == 'A3')
```

There are two conditions here. Let's filter for data from the second condition only (`Condition == Two`).

```{r}
A3_second <- A3 %>% filter(Condition == 'Two')
```

Now, we will use the function `lm()` to compute $m$ (slope) and $b$ (y-intercept).

In R, the relationship between $y$ (dependent variable) and $x$ (independent variable) is written as `y~x` using tilde (`~`). In other words, instead of directly writing $y = mx + b$ in R, we use `~` to describe the relationship between $y$ and $x$. Let's write the relationship between `absBP` (dependent variable) and `logSF` (independent variable) within the function `lm()`.

```{r}
lm(df$absBP ~ df$logSF)
```

This yields two main outputs. Let's store this result using a new variable `res`, which is short for **results**.

```{r}
res <- lm(A3_second$absBP ~ A3_second$logSF)
summary(res)
```

`summary()` is a function that prints all the outputs from a given model object. Here, the model object is `res`, which has been created using the `lm()` function. When using `lm()`, it is advisable to always store the results and print the results using `summary()`. For more information about `summary()`, please check out `?summary`.

We see that the y-intercept $b$ is 0.4582 under the column `Estimate`. We also see that the slope $m$ is 0.3046 under the column `Estimate`. You can ignore all the other values for now. 

Let's visualize the data of A3 and fit a linear slope.

```{r, fig.width = 3.65, fig.height = 3.65}
A3_second %>% ggplot(aes(x = logSF, y = absBP)) +
  geom_point() +
  geom_abline(slope = 0.4582, intercept = 0.3046)
```

Instead of writing the slope and intercept manually, we can subset these values using `$`.

```{r}
res$coefficients
```

The first value is the intercept. Therefore use `[[1]]` to subset the intercept.

```{r}
res$coefficients[[1]]
```

You can also subset the slope using `[[2]]`. 

```{r}
res$coefficients[[2]]
```

Now let's plot the graph again and label the y-intercept as well.
A separate data frame `y_int` containing the y-intercept is created below. The code below is a bit challenging, so please use `?` to figure out each function if you are not sure. 

```{r, fig.width = 3.65, fig.height = 3.65}
y_int <- data.frame(logSF = 0, absBP = res$coefficients[[1]])

A3_second %>% ggplot(aes(x = logSF, y = absBP)) +
  geom_point(size = 3) +
  geom_abline(slope = res$coefficients[[2]], 
              intercept = res$coefficients[[1]]) +
  geom_point(data = y_int, color = sm_color('red'), size = 3) +
  sm_corr_theme() +
  annotate('text', x = 0, y = 1.2, size = 3.5,
           label = paste('Slope =',round(res$coefficients[[2]],2))) +
  annotate('text', x = 0, y = 0.9, size = 3.5,
           label = paste('Intercept =', round(res$coefficients[[1]],2)))
```

In summary, you can compute the slope of a linear function between $y$ and $x$ and using `lm()`, where you use `~` to describe the relationship. You also need to use `$` to subset the slopes directly. 

## Calculating slopes of all subjects, groups and conditions

Now that we know how to compute slope for each subject, let's calculate slopes of our entire dataset.

```{r}
df
```

We see that there are 20 subjects total, each of which has completed two conditions. So there are 40 slopes to calculate! Does that mean we need to use `lm()` 40 times?

The answer is **no**. **smplot** has a function `sm_slope_list()` that returns a dataframe of slopes from **linear regression**. It works similarly to `sm_auc_list()`.

- `data` = this argument requires the variable that stores the data frame. In our case, it is `df`.
- `subjects` = this argument requires the name of the column of the data frame that contains subjects. It must strings, ex. `'Subject'`, not `Subject`.
- `groups` = this argument requires the name of the column of the data frame that contains each group. In this example, there is no group. An example would be `Group` column that contains two groups: `Normal` and `Amblyopia`.
- `conditions` = this argument requires name of the column of the data frame that contains each condition. In our example, the two conditions are `One` and `Two`.
- `x` = this argument requires the name of the column of the data frame that contains the x-axis points from which the AUC can be calculated. In our case, these are values from the `logSF` column of `df`. It must  strings, ex. `'logSF'`, not `logSF`.
- `values` = this argument requires the name of the column of the data frame that contains the actual data, which are the y-axis points from which the AUC can be calculated. In our case, it is the change in contrast balance ratio. It must strings, ex. `'absBP'`, not `absBP`.

We can store the results from `sm_slope_list()` into a new variable. I will call the new variable `slope_df`.

```{r}
slope_df <- sm_slope_list(subjects = 'Subject', 
                          conditions = 'Condition',
                          groups = 'Group',
                          x = 'logSF', values = 'absBP', 
                          data = df)

slope_df
```

We see that the slope of A3 in `Two` condition is identical to the one we have obtained from the `lm()` function.

```{r}
res$coefficients[[2]] 
```

```{r}
slope_df %>% filter(Subject == 'A3' & Condition == 'Two') %>%
  select(Slope)
```

Now we can plot all the slopes from `One` condition using `sm_bar()`, `sm_boxplot` etc. Let's have a try.

The factor level of `slope_df_one$Group` is changed below so that bar plot of normal observer is plotted first, then amblyopia (rather than amblyopia -> normals as per alphabetical order).


```{r}
slope_df_one <- slope_df %>% filter(Condition == 'One')
slope_df_one$Group <- factor(slope_df_one$Group, 
                             levels = c('Normal','Amblyopia'))
```


Here is a boxplot showing `slope_df_one`'s data. 

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
slope_df_one %>% ggplot(mapping = aes(x = Group, y = Slope, color = Group)) +
  sm_boxplot(alpha = 0.6) + 
  scale_color_manual(values = sm_palette(2)) +
  ggtitle('\u0394 AUC re: Each day\'s baseline')
```

Here is a bar graph showing `slope_df`'s data for `One` condition only. This figure is similar to Figure 3B in the paper (Mao et al., 2020).

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
slope_df_one %>% ggplot(aes(x = Group, y = Slope, fill = Group)) +
   sm_bar(shape = 21, color = 'white', 
          bar_fill_color = 'gray80',
          point_alpha = 1) +
  scale_fill_manual(values = sm_palette(2)) +
  ggtitle('Binocular imbalance')
```
