---
output: html_document
chapter: Plotting the Contrast Sensitivity Function
editor_options:
  chunk_output_type: console
---

# Plotting the Contrast Sensitivity Function (in Progress)

In this section, we will plot the contrast sensitivity function using the **smCSF** package. I will complement it with **smplot** for aesthetics. 

We will use data of achromatic contrast sensitivity from 51 normal observers from this paper (a subset of the entire dataset):

**Kim, Y. J., Reynaud, A., Hess, R. F., & Mullen, K. T. (2017). A normative data set for the clinical assessment of achromatic and chromatic contrast sensitivity using a qCSF approach. Investigative ophthalmology & visual science, 58(9), 3628-3636.**

Luckily, the data are publicly available. I have reorganized the data so that we could easily use them in the form of a data frame. Let's load the data from online.


```{r}
library(tidyverse)
library(smplot)
library(smCSF)
```

```{r, eval = F}
ACh <- read_csv('https://www.smin95.com/data_ACh.csv')
```

```{r, echo = F}
ACh <- read_csv('data_Ach.csv')
```

```{r}
head(ACh)
```

There are four columns in this data frame:

- First, `Subject` refers to each participant. There are 51 participants total in the .csv file.

- Next, `Repetition` refers to each repetition of the measurement The participants performed two repetitions of the contrast sensitivity measurement.

- `SpatialFreq` refers to spatial frequency of the stimuli that were shown to test the observers' contrast sensitivity. There are a total of 12 spatial frequencies.

- The `Sensitivity` column refers to the linear values of the contrast sensitivity.

## Brief Introduction

The contrast sensitivity function (CSF) how the observer is sensitive to an image. This image could be large (low spatial frequency) or small (high spatal frequency). To obtain the CSF, the measurement test must measure the threshold. Threshold refers to the difference in stimulus magnitude between the states of presence and absence. If the observer can just barely discriminate between these two states of the visual stimulus, this can be described as the threshold for detection. 

## Data Visualization

First, let's compute the average, standard errors and standard deviation using the dataset. It is recommended that the reader has read Chapters 5 and 6 already. Understanding of `summarise()`, `filter()`, `mutate()` and `%>%` is assumed.

The code below reorganizes the `ACh` dataset so that instead of containing the data of each subject, it now contains the averaged data across `Repetition` and `SpatialFreq` as well as their standard errors and standard deviations. We use `mean()` to compute the mean, `sm_stdErr()` the standard error, and `sd()` the standard deviation.

```{r}
ACh_avg <- ACh %>% 
  group_by(Repetition, SpatialFreq) %>%
  summarise(avgSens = mean(Sensitivity),
            stdErrSens = sm_stdErr(Sensitivity),
            stdDevSens = sd(Sensitivity))
  
ACh_avg
```

Let's create a variable that stores the data for each `Repetition`. Before we do that, however, it is important that we convert the data type of the `Repetition` column from `<dbl>` to `<fct>`, which refers from **double** to **factor**. **Double** refers to continuous data, whereas **factor** refers to categorical data. 

```{r}
ACh_avg$Repetition <- factor(ACh_avg$Repetition) # factor

ACh_avg1 <- ACh_avg %>% filter(Repetition == 1) # repetition 1
ACh_avg2 <- ACh_avg %>% filter(Repetition == 2) # repetition 2
```

Let's now plot the data for each repetition using **ggplot2**. Understanding of Chapters 1-4 is assumed. 

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  geom_point() +
  sm_hgrid() +
  xlab('Spatial frequency (c/deg)') +
  ylab('Averaged sensitivity') +
  ggtitle('Repetition 1 (n=51)')
```

So far, this graph does not show the contrast sensitivity function, which has a rotund shape. Instead, if we connect the points, we get this plot instead.

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  geom_point() +
  geom_line() +
  sm_hgrid() +
  xlab('Spatial frequency (c/deg)') +
  ylab('Averaged sensitivity') +
  ggtitle('Repetition 1 (n=51)')
```

The function is still not a smooth, rotund shape. This is because it is standard to plot the contrast sensitivity function in log10 scales in both x and y axes. The code below shows the plot in the log scale.

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  geom_point() +
  geom_line() +
  sm_hgrid() +
  xlab('Spatial frequency (c/deg)') +
  ylab('Averaged sensitivity') +
  ggtitle('Repetition 1 (n=51)') +
  scale_x_continuous(trans = 'log10') +
  scale_y_continuous(trans = 'log10')
```

This plot shows the averaged sensitivity for each spatial frequency during the first repetition. However, the contrast sensitivity function is still missing because this plot instead shows the plot that is connected between a pair of the points. To plot the contrast sensitivity function with the data, `sm_CSF()` is required.

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_CSF()
```

Notice that even if the data frame contains data in **linear units**, the plotted data from `sm_CSF()` are in log10 scales. It shows a nice, rotund shape that is typical of the contrast sensitivity function. The function can also plot the linear curve by setting `sm_CSF(logXY = FALSE)`. The default is `sm_CSF(logXY = TRUE)`.

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_CSF(logXY = FALSE)
```

Now let's plot a prettier contrast sensitivity function.

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_CSF(color = sm_color('wine')) +
  geom_point(color = sm_color('darkred'), size = 3) +
  sm_hgrid() +
  xlab('Spatial frequency (c/deg)') +
  ylab('Averaged sensitivity') +
  ggtitle('Repetition 1 (n=51)') 
```

We can also plot the data using `ACh_avg2`, which contains data from the second repetition rather than first.

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
ACh_avg2 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_CSF(color = sm_color('lightgreen')) +
  geom_point(color = sm_color('viridian'), size = 3) +
  sm_hgrid() +
  xlab('Spatial frequency (c/deg)') +
  ylab('Averaged sensitivity') +
  ggtitle('Repetition 2 (n=51)') 
```

Notice that `sm_CSF()` is written first because the curve has to appear before the points (`geom_point()`); else, the points will be covered by the CSF curve. This example is shown below.

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
ACh_avg2 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  geom_point(color = sm_color('viridian'), size = 3) +
   sm_CSF(color = sm_color('lightgreen')) +
  sm_hgrid() +
  xlab('Spatial frequency (c/deg)') +
  ylab('Averaged sensitivity') +
  ggtitle('Repetition 2 (n=51)') 
```

The curve appears later in this plot, so it covers some overlapping points. 

**smCSF** also offers a function that fill the area of the CSF curve. This function is `sm_areaCSF()`. The default is set to plot the area in log10 scales, `sm_areaCSF(logXY = TRUE)`.

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_areaCSF() +
  sm_CSF() 
```

To plot the linear CSF and its area, `logXY = FALSE` has to be set `FALSE` in both `sm_areaCSF()` and `sm_CSF()`. The functions of `smCSF` have defaults `logXY = TRUE`, therefore, it is recommended that the user uploads a data frame that contains the linear data of spatial frequencies and contrast sensitivities. The functions then will convert them into log scale by default.

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_areaCSF(logXY = FALSE) +
  sm_CSF(logXY = FALSE) 
```

We can make the area plot prettier.

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
ACh_avg2 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_areaCSF(fill = sm_color('lightgreen'), alpha = 0.5) +
  sm_CSF(color = sm_color('viridian')) +
  sm_hgrid() +
  xlab('Spatial frequency (c/deg)') +
  ylab('Averaged sensitivity') +
  ggtitle('Repetition 2 (n=51)') +
  scale_y_log10(limits = c(1,100)) +
  annotate('text', x = 0.9, y = 1.3, label = 'Area under Log CSF')
```

You might get a warning that says `Scale for 'y' is already present. Adding another scale for 'y', which will replace the existing scale`. You can ignore this warning and not be worried about it. This is because we used the function `scale_y_log10()` on top of `sm_CSF()`, which actually calls forth the function `scale_y_continuous()` to convert the plot into log scales. Hence, there are two functions that serve the same purpose here, so the warning message is printed.

The **smCSF** package also offers a function that draws a ribbon shade to denote standard error or standard deviation. This function works exactly the same as `geom_ribbon()`. It requires the uses to specify `ymin` and `ymax`. The example below plots standard error using `sm_ribbonCSF()`. As before, its default is `logXY = TRUE`. Therefore, even if `ACh_avg2` contains data in linear units, the ribbon shade that is generated by `sm_ribbonCSF()` will be in log unit. 

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
ACh_avg2 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_ribbonCSF(aes(ymin = avgSens - stdErrSens,
                   ymax = avgSens + stdErrSens)) +
  sm_CSF() 
```

Let's make this plot prettier using functions from **smplot**.

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
ACh_avg2 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_ribbonCSF(aes(ymin = avgSens - stdErrSens,
                   ymax = avgSens + stdErrSens),
               fill = sm_color('lightgreen'), # fill controls the color of the shade
               alpha = 0.5) +
  sm_CSF(color = sm_color('viridian')) +
  xlab('Spatial frequency (c/deg)') +
  ylab('Averaged sensitivity') +
  ggtitle('Repetition 2 (n=51)') +
  sm_hgrid()
```

You can also shade standard deviation instead, where `ymin = avgSens - stdDevSens` and `ymax = avgSens + stdDevSens`.

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_ribbonCSF(aes(ymin = avgSens - stdDevSens,
                   ymax = avgSens + stdDevSens)) +
  sm_CSF() 
```

Let's make the plot prettier using **smplot**.

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
ACh_avg1 %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_ribbonCSF(aes(ymin = avgSens - stdDevSens,
                   ymax = avgSens + stdDevSens),
               fill = sm_color('wine'), alpha = 0.4) +
  sm_CSF(color = sm_color('darkred')) +
  xlab('Spatial frequency (c/deg)') +
  ylab('Averaged sensitivity') +
  ggtitle('Repetition 2 (n=51)') +
  sm_hgrid()
```

Unfortunately, in this case, the standard deviation is quite broad. Let's plot the averaged data across all repetitions and see if the standard deviation is still too broad.

```{r}
ACh_avg_final <- ACh %>% group_by(SpatialFreq) %>%
  summarise(avgSens = mean(Sensitivity),
            stdErrSens = sm_stdErr(Sensitivity),
            stdDevSens = sd(Sensitivity))

head(ACh_avg_final)
```

Here is the new data frame `ACh_avg_final`.

```{r, fig.width = 3.65, fig.height = 3.65, warning = F}
ACh_avg_final %>% ggplot(aes(x = SpatialFreq, y = avgSens)) +
  sm_ribbonCSF(aes(ymin = avgSens - stdDevSens,
                   ymax = avgSens + stdDevSens),
               fill = sm_color('skyblue'), alpha = 0.25) +
  sm_CSF(color = sm_color('blue'),
         linetype = 'dashed') +
  xlab('Spatial frequency (c/deg)') +
  ylab('Averaged sensitivity') +
  ggtitle('Contrast Sensitivity (n=51)') +
  sm_hgrid() +
  geom_point(color = sm_color('blue'), size = 3) 
```


The standard deviation seems to be still broad.

## Facetting the Contrast Sensitivity Functions

To facet means to split the plot into multiple plots based on a categorical variable (i.e., **factor**). We will facet the contrast sensitivity function plot above (the blue curve) into two panels based on two repetitions (red and green below).

```{r, fig.width = 6.8, fig.height = 3.65, warning = F, echo = F}
ACh_avg %>%
  ggplot(aes(x = SpatialFreq, y = avgSens,
             group = Repetition, color = Repetition,
             fill = Repetition)) +
   sm_ribbonCSF(aes(ymin = avgSens - stdDevSens,
                   ymax = avgSens + stdDevSens), alpha = 0.4) +
   sm_CSF(linetype = 'dashed') +
   sm_hgrid() +
   sm_facet_header(font_size=11) +
   geom_point(size = 3) +
   scale_fill_manual(values = sm_color('wine','lightgreen')) +
   scale_color_manual(values = sm_color('darkred','viridian')) +
   xlab('Spatial frequency (c/deg)') +
   ylab('Averaged sensitivity') +
   facet_wrap(~ Repetition,
             labeller = as_labeller(c('1' = 'Repetition 1',
                                      '2' = 'Repetition 2')))  
```

To facet the plot, you need to have a data frame that has data with a column that specifies the categorical variable. For instance, let's look at the data frame `ACh_avg`.

```{r}
ACh_avg
```

We see that the first column of `Ach_avg` has a column named `Repetition` which stores whether the given data are from the first or second repetition. To see all the unique levels of `Repetition`, we can use the function `unique()`.

```{r}
unique(ACh_avg$Repetition)
```

To facet a plot using ggplot2, the function `facet_wrap()` has to be used.

```{r, fig.width = 6.8, fig.height = 3.65, warning = F}
ACh_avg %>%
  ggplot(aes(x = SpatialFreq, y = avgSens)) +
   sm_CSF() +
   facet_wrap(~ Repetition) 
```

The argument inside `facet_wrap` has to include the tilde (`~`). The right side the tilde (`~`) should be the name of the categorical variable by which you wish to facet the plot. If the variable however happens to be continuous, `facet_wrap()` would not work properly.

```{r, fig.width = 6.8, fig.height = 3.65, warning = F}
ACh_avg %>%
  ggplot(aes(x = SpatialFreq, y = avgSens)) +
   sm_CSF() +
   facet_wrap(~ stdErrSens) 
```

However, I do not often use `facet_wrap()` because the plot often ends up not being pretty. Let's try to make it prettier.

```{r, fig.width = 6.8, fig.height = 3.65, warning = F}
ACh_avg %>%
  ggplot(aes(x = SpatialFreq, y = avgSens,
             group = Repetition, color = Repetition,
             fill = Repetition)) +
   sm_ribbonCSF(aes(ymin = avgSens - stdDevSens,
                   ymax = avgSens + stdDevSens), alpha = 0.4) +
   sm_CSF(linetype = 'dashed') +
   sm_hgrid() +
   geom_point(size = 3) +
   scale_fill_manual(values = sm_color('wine','lightgreen')) +
   scale_color_manual(values = sm_color('darkred','viridian')) +
   xlab('Spatial frequency (c/deg)') +
   ylab('Averaged sensitivity') +
   facet_wrap(~ Repetition) 
```

Personally, I do not like the grey-filled header of each panel. Since I personally will find myself repeatedly bypassing the default grey color into a color of my choice, I have created the function myself `sm_facet_header()` as part of the **smCSF** package.

```{r, fig.width = 6.8, fig.height = 3.65, warning = F}
ACh_avg %>%
  ggplot(aes(x = SpatialFreq, y = avgSens,
             group = Repetition, color = Repetition,
             fill = Repetition)) +
   sm_ribbonCSF(aes(ymin = avgSens - stdDevSens,
                   ymax = avgSens + stdDevSens), alpha = 0.4) +
   sm_CSF(linetype = 'dashed') +
   sm_hgrid() +
   geom_point(size = 3) +
   scale_fill_manual(values = sm_color('wine','lightgreen')) +
   scale_color_manual(values = sm_color('darkred','viridian')) +
   xlab('Spatial frequency (c/deg)') +
   ylab('Averaged sensitivity') +
   facet_wrap(~ Repetition) +
   sm_facet_header()
```

Now the title of each plot has been bolded and the filled color is white. I personally think that this plot is a lot cleaner than the default one.

Next, I wish to change the titles of each panel because `1` and `2` are not descriptive. To do so, we need to use the argument `labeller` within `facet_wrap()` to specify the title of each panel.

```{r, fig.width = 6.8, fig.height = 3.65, warning = F}
ACh_avg %>%
  ggplot(aes(x = SpatialFreq, y = avgSens,
             group = Repetition, color = Repetition,
             fill = Repetition)) +
   sm_ribbonCSF(aes(ymin = avgSens - stdDevSens,
                   ymax = avgSens + stdDevSens), alpha = 0.4) +
   sm_CSF(linetype = 'dashed') +
   sm_hgrid() +
   sm_facet_header(font_size=11) +
   geom_point(size = 3) +
   scale_fill_manual(values = sm_color('wine','lightgreen')) +
   scale_color_manual(values = sm_color('darkred','viridian')) +
   xlab('Spatial frequency (c/deg)') +
   ylab('Averaged sensitivity') +
   facet_wrap(~ Repetition,
             labeller = as_labeller(c('1' = 'Repetition 1',
                                      '2' = 'Repetition 2'))) 
```

We use the function `as_labeller()` as part of the argument `labeller()` inside `facet_wrap()` to specify that factor level `1` should be shown as `Repetition 1` and `2` as `Repetition 2`.

We can also change the number of rows and columns using `facet_wrap()`. Lets make a data frame that has 4 repetitions instead of 2. 

```{r}
set.seed(1)
spatFreqNum <- length(unique(ACh_avg_4rep$SpatialFreq))
ACh_avg_4rep <- rbind(ACh_avg,ACh_avg)
ACh_avg_4rep$Repetition <- rep(1:4,each=spatFreqNum)
ACh_avg_4rep[25:nrow(ACh_avg_4rep), 3:5] <- ACh_avg_4rep[25:nrow(ACh_avg_4rep), 3:5] + round(rnorm(24),1) 
ACh_avg_4rep$Repetition <- factor(ACh_avg_4rep$Repetition)
```

We can specify the number of columns (ex. `ncol = 2`) and rows (ex. `nrow = 2`) inside the function `facet_wrap()`, as shown in the example below.

```{r, fig.width = 6.8, fig.height = 6.8, warning = F}
ACh_avg_4rep %>%
  ggplot(aes(x = SpatialFreq, y = avgSens,
             group = Repetition, color = Repetition,
             fill = Repetition,
             alpha = Repetition)) +
   sm_ribbonCSF(aes(ymin = avgSens - stdDevSens,
                   ymax = avgSens + stdDevSens)) +
   sm_CSF(linetype = 'dashed') +
   sm_hgrid() +
   sm_facet_header(font_size=11) +
   geom_point(size = 3, alpha = 1) +
   scale_fill_manual(values = sm_color('wine','lightgreen','yelloworange', 'pink')) +
   scale_color_manual(values = sm_color('darkred','viridian', 'orange','crimson')) +
   scale_alpha_manual(values = c(0.4,0.4,0.3,0.2)) +
   xlab('Spatial frequency (c/deg)') +
   ylab('Averaged sensitivity') +
   facet_wrap(~ Repetition, nrow=2, ncol=2,
             labeller = as_labeller(c('1' = 'Repetition 1',
                                      '2' = 'Repetition 2',
                                      '3' = 'Repetition 3',
                                      '4' = 'Repetition 4'))) 
```

Notice that when `facet_wrap()` is used, there is no need to create variable for each plot like we have done in Chapter 5 and use `sm_common_axis()`. The preference can vary depending on the user. I still prefer to store each plot so that I can have freedom to decide where to place each panel wherever I want. 

In the next chapter, we will discuss on some techniques to analyze the data of contrast sensitivity.