---
output: html_document
chapter: Raincloud and Forest Plots
editor_options:
  chunk_output_type: console
---

# Raincloud and Forest Plots

Please load these three packages to go through the chapter.

```{r, warning = FALSE, message = FALSE}
library(tidyverse) 
library(cowplot) 
library(smplot2)
```

## Raincloud plot

A raincloud plot shows the raw data, probability density and summary statistics such as the median, quantiles etc by presenting individual data, a violin plot and a boxplot together.

It might be appealing to present all these components simultaneously to ensure that the data are shown transparently. However, whether this is truly necessary in the field of psychological/neuroscience research has not been entirely resolved because showing all at the same time can be visually overwhelming.

Nevertheless, I have decided to create a function `sm_raincloud()`. It is flexible and enables users to fine-control the aesthetics of the raincloud plot. I will leave it up to users to decide whether using raincloud is appropriate. Most importantly, this can depend on the type and nature of data set that is used.

Let's generate random data to explore this function.

```{r}
set.seed(211) # generate random data
group1 = abs(rnorm(20,2.5,1))
group2 = abs(rnorm(20,5,1))
Subject <- rep(paste0('S',seq(1:20)), 2)
Data <- data.frame(Value = matrix(c(group1,group2),ncol=1))
Group <- rep(c('Group 1', 'Group 2'), each = length(group1))
df <- cbind(Subject, Data, Group)
```

In this data set, there are two distinct groups.

```{r, fig.width = 3.5, fig.height=3.5}

ggplot(data = df, mapping = aes(x = Group, y = Value, fill = Group)) +
  sm_raincloud() +
  scale_fill_manual(values = sm_color('blue','orange'))

```

The default of the individual points' `shape` is 21, which is circle with borders. Therefore `fill` parameter has to be specified to change the filling colors. 

After removing the defaults from **smplot2** by including `xxx.param = list()`, you can set the color of all visual components by setting `color = 'black'` as a generic parameter. Note that the boxplot and violin have both `fill` and `color` parameters. When `color = 'black'` is included as a generic parameter, `shape` of the point becomes 16 because the default of *smplot2* is overridden. Therefore, now that the point has no border, `color` affects the filling color rather than `fill`.

```{r, fig.width = 3.5, fig.height=3.5}
ggplot(data = df, mapping = aes(x = Group, y = Value, color = Group)) +
  sm_raincloud(color = 'black', point.params = list(),
               boxplot.params = list(),
               violin.params = list()) 
```

The limitation of this function is that you cannot assign unique to the boxplot for each group using `scale_color_manual()`. We can test this using the script below.

```{r, fig.width = 3.5, fig.height=3.5}

ggplot(data = df, mapping = aes(x = Group, y = Value, color = Group)) +
  sm_raincloud(point.params = list(),
               boxplot.params = list(),
               violin.params = list()) +
  scale_color_manual(values = sm_color('blue','orange')) 

```

In this example, the `color` affects the border color of the violin. However, the boxplot's line color is still black even with `scale_color_manual()`.

```{r, fig.width = 3.5, fig.height=3.5}

ggplot(data = df, mapping = aes(x = Group, y = Value, fill = Group)) +
  sm_raincloud(point.params = list(),
               boxplot.params = list(),
               violin.params = list()) +
  scale_fill_manual(values = sm_color('blue','orange')) 

```

In this example, the `fill` affects the filling color of the violin but not points. 

In most cases, users might not need to manually tweak the defaults of *smplot2*. This is the default appearance with two custom colors `darkred` and `viridian`.

```{r, fig.width = 3.5, fig.height=3.5}

ggplot(data = df, mapping = aes(x = Group, y = Value, fill = Group)) +
  sm_raincloud() +
  scale_fill_manual(values = sm_color('darkred','viridian'))

```

Notice that there is an outlier in Group 2. The black circle at the top that is aligned to the top `viridian` point indicates that the `viridian` point is an outlier. This option of indicating outlier can be removed setting `boxplot.params = list(outlier.shape = NA)`.

```{r, fig.width = 3.5, fig.height=3.5}

ggplot(data = df, mapping = aes(x = Group, y = Value, fill = Group)) +
  sm_raincloud(boxplot.params = list(outlier.shape = NA)) +
  scale_fill_manual(values = sm_color('darkred','viridian'))

```

To truly modify each component of the raincloud plot that is generated using `sm_raincloud()`, you have to understand these `xxx.param = list()`:

- `boxplot.params` feeds arguments to `geom_half_boxplot()` (almost identical to `geom_boxplot()`) to plot the boxplot with specified `color`, `fill`, `outlier.shape` etc.

- `violin.params` feeds arguments to `geom_half_violin()` (almost identical to `geom_violin()`) to plot the violin plot with specified `color`, `alpha` and `fill`, etc

- `point.params` feeds arguments to `geom_point()` to plot the points in the shadow with specified `color`, `fill`, `size`, `shape` and `alpha`. 

### Changing the x-axis labels

Note that the x-axis title is `Group` and the x-axis tick also says `Group 1` and `Group 2`, making the labels repetitive. Therefore we can consider to modify the tick labels as `One` and `Two` using `scale_x_discrete()`.

```{r, fig.width = 3.5, fig.height=3.5}

ggplot(data = df, mapping = aes(x = Group, y = Value, fill = Group)) +
  sm_raincloud(boxplot.params = list(outlier.shape = NA)) +
  scale_fill_manual(values = sm_color('darkred','viridian')) +
  scale_x_discrete(labels = c('One', 'Two'))

```

Note that in **smplot**, `scale_x_continous()` was used to modify the label at each x level. In **smplot2**, `scale_x_discrete()` is used instead because `sm_raincloud()` keeps the x-axis `discrete` rather than forcibly converting it to becoming `continuous`.

`scale_x_discrete()` is used here because the data frame that is used to plot the data has categorical x variables as shown below.

```{r}
df$Group
```

This shows that `df$Group` is discrete and has two unique levels.

```{r}
unique(df$Group)
```

Instead of changing the x-tick label, we can remove the x-axis title instead by writing `theme(axis.title.x = element_blank())`.

```{r, fig.width = 3.5, fig.height=3.5}

ggplot(data = df, mapping = aes(x = Group, y = Value, fill = Group)) +
  sm_raincloud(boxplot.params = list(outlier.shape = NA)) +
  scale_fill_manual(values = sm_color('darkred','viridian')) +
  theme(axis.title.x = element_blank())

```

### Separating the components and configuration

`sm_raincloud()` has an interesting parameter `sep_level`. It enables users to separate components of the raincloud plot. It is a reduced word of separation level and has a range between 0 and 4. 

When `sep_level = 0`, all components are overlapped. If `sep_level = 4`, all components are separated. The default of `sm_raincloud()` is set to `sep_level = 2`.

```{r, fig.width = 3.5, fig.height=3.5}

ggplot(data = df, mapping = aes(x = Group, y = Value, fill = Group)) +
  sm_raincloud(boxplot.params = list(outlier.shape = NA),
               sep_level = 0) +
  scale_fill_manual(values = sm_color('darkred','viridian')) +
  theme(axis.title.x = element_blank())

```

```{r, fig.width = 3.5, fig.height=3.5}

ggplot(data = df, mapping = aes(x = Group, y = Value, fill = Group)) +
  sm_raincloud(boxplot.params = list(outlier.shape = NA),
               sep_level = 4) +
  scale_fill_manual(values = sm_color('darkred','viridian')) +
  theme(axis.title.x = element_blank())

```

You also change the configuration of the raincloud plot so that it faces to the left rather than to the right using the argument `which_side`.

- `which_side` controls the configuration of the raincloud plot. It accepts `'r'` and `'right'` for right configuration (violin plot facing to the right), and `'l'` and `'left'` for left configuration.

```{r, fig.width = 3.5, fig.height=3.5}

ggplot(data = df, mapping = aes(x = Group, y = Value, fill = Group)) +
  sm_raincloud(boxplot.params = list(outlier.shape = NA),
               sep_level = 0,
               which_side = 'l') +
  scale_fill_manual(values = sm_color('darkred','viridian')) +
  theme(axis.title.x = element_blank())

```

### Changng the theme and orientation

`sm_raincloud()` uses `sm_hgrid()` automatically and receives `legends` and `borders` as arguments as all other visualizing functions of **smplot2**.

```{r, fig.width = 3.5, fig.height=3.5}

ggplot(data = df, mapping = aes(x = Group, y = Value, fill = Group)) +
  sm_raincloud() +
  scale_fill_manual(values = sm_color('darkred','viridian')) +
  theme(axis.title.x = element_blank())
```

We can remove the borders by setting `borders = FALSE`.

```{r, fig.width = 3.5, fig.height=3.5}

ggplot(data = df, mapping = aes(x = Group, y = Value, fill = Group)) +
  sm_raincloud(borders = F) +
  scale_fill_manual(values = sm_color('darkred','viridian')) +
  theme(axis.title.x = element_blank()) 
```

Also, you can bypass `sm_hgrid()` by adding another theme, such as `sm_minimal()`.

```{r, fig.width = 3.5, fig.height=3.5}

ggplot(data = df, mapping = aes(x = Group, y = Value, fill = Group)) +
  sm_raincloud() +
  scale_fill_manual(values = sm_color('darkred','viridian')) +
  theme(axis.title.x = element_blank()) +
  sm_minimal()
```

You can also change the orientation of the raincloud plot from vertical to horizontal by setting `vertical = FALSE`.

```{r, fig.width = 3.5, fig.height=3.5}

ggplot(data = df, mapping = aes(x = Group, y = Value, fill = Group)) +
  sm_raincloud(vertical = FALSE) +
  scale_fill_manual(values = sm_color('darkred','viridian')) 
```

We can remove the y-axis title of the plot by using `theme(axis.title.y = element_blank())` instead ` theme(axis.title.x = element_blank())`.

```{r, fig.width = 3.5, fig.height=3.5}

ggplot(data = df, mapping = aes(x = Group, y = Value, fill = Group)) +
  sm_raincloud(vertical = FALSE) +
  scale_fill_manual(values = sm_color('darkred','viridian')) +
  theme(axis.title.y = element_blank()) 
```


### Raincloud plot with data from multiple groups

`sm_raincloud()` can also raincloud plots for data from multiple groups. Let's load this fake data set.

```{r, message = F, warning = F}
df2 <- read_csv('https://www.smin95.com/amblyopia_random.csv')
df2a <- filter(df2, Condition == 'One')  
df2a <- filter(df2, SF %in% c(0.5,8))
df2a$SF <- as.factor(df2a$SF)
```

Make sure `SF` is converted to `factor` first because the x-scale has to be **discrete**, not continuous. In this example, we have filtered for `SF` that equals `0.5` or `8`.

```{r, fig.width = 3.5, fig.height=3.5}

ggplot(data = df2a, mapping = aes(x = SF, y = absBP, fill = Group,
                                 color = Group)) +
  sm_raincloud(sep_level = 2) +
  scale_fill_manual(values = sm_color('darkred','viridian')) 
```

We immediately see that the plot is extremely crowded because there are four violin plots, four boxplots and overlapping individual points. 

We could still try to make it visually neat, however. In the upcoming approach, we will use two separate data frames, one for each group (`Normal` and `Amblyopia`).

```{r}
df2amb <-  filter(df2a, Group == 'Amblyopia')
df2norm <-  filter(df2a, Group == 'Normal')
```

Next we first plot the data from one of the data frames. Make sure that the first set of raincloud plot is faced slightly to the right using `position_nudge()`.

```{r, fig.width = 3.5, fig.height=3.5, warning = FALSE}
ggplot(mapping = aes(x = SF, y = absBP, fill = Group)) +
  sm_raincloud(data = df2amb, position = position_nudge(x = +0.3)) +
  scale_fill_manual(values = sm_color('darkred')) 
```

Next, we plot the `viridian` (`Normal`) data on the figure which `left` configuration.

```{r, fig.width = 3.5, fig.height=3.5, warning = FALSE}
ggplot(mapping = aes(x = SF, y = absBP, fill = Group)) +
  sm_raincloud(data = df2amb, position = position_nudge(x = +0.15),
               point.params = list(size = 3, shape = 21,
                                   color = 'transparent', alpha = 0.4,
                                   position = sdamr::position_jitternudge(nudge.x = 0.06, 
                                                                   jitter.width = 0.06))) +
  scale_fill_manual(values = sm_color('darkred', 'viridian')) +
  sm_raincloud(data = df2norm, which_side = 'left',
               position = position_nudge(x = -0.15),
               point.params = list(size = 3, shape = 21,
                                   color = 'transparent', alpha = 0.4,
                                   position = sdamr::position_jitternudge(nudge.x = -0.06, 
                                                                   jitter.width = 0.06)))
```

I think this attempt is somewhat acceptable. Let's improve the aesthetics slightly by adjusting the `size` of the point and adding `legends`.

```{r, fig.width = 3.5, fig.height=3.5, warning = FALSE}
ggplot(mapping = aes(x = SF, y = absBP, fill = Group)) +
  sm_raincloud(data = df2amb, position = position_nudge(x = +0.15),
               show.legend = FALSE,
               point.params = list(size = 2, shape = 21,
                                   color = 'transparent', 
                                   show.legend = TRUE,
                                   alpha = 0.3,
                                   position = sdamr::position_jitternudge(nudge.x = 0.06, seed = 10,
                                                                   jitter.width = 0.06))) +
  scale_fill_manual(values = sm_color('darkred', 'viridian')) +
  sm_raincloud(data = df2norm, which_side = 'left',
               show.legend = FALSE,
               position = position_nudge(x = -0.15),
               point.params = list(size = 2, shape = 21,
                                   show.legend = TRUE,
                                   color = 'transparent', alpha = 0.3,
                                   position = sdamr::position_jitternudge(nudge.x = -0.06, seed = 10,
                                                                   jitter.width = 0.06))) +
  theme(legend.position = c(0.24,0.85),
        legend.title = element_blank(),
        legend.text = element_text(size = 11)) +
  ggtitle('Raincloud plot of two groups') +
  scale_y_continuous(limits = c(0,2.5)) 
```

Another solution is to plot the raincloud plot for each group per panel, and put two panels together. In fact, I don't recommend the reader to plot data of multiple groups within the same figure when drawing a raincloud plot.

**Exercise**: By using this script above with some minor changes, you can also generate boxplot (`sm_boxplot()`) and violin (`sm_violin()`) using data from multiple groups. You can try them on your own.

## Forest plot

A forest plot is commonly known as a visualization method to summarise data across multiple studies or experiments. For this reason, it has been widely used in meta analysis studies that aim to summarise the results of tens or hundreds of studies into a few figures. I believe that forest plot's usage can exceed beyond simple meta analysis, all the way to describing results from a single study with numerous experimental conditions.

Let's generate some random data first and draw a forest plot using `sm_forest()` to explore what the function provides.

```{r}
set.seed(2) # generate random data
day1 = rnorm(20,0,1)
day2 = rnorm(20,5,1)
day3 = rnorm(20,6,1.5)
day4 = rnorm(20,7,2)
Subject <- rep(paste0('S',seq(1:20)), 4)
Data <- data.frame(Value = matrix(c(day1,day2,day3,day4),ncol=1))
Day <- rep(c('Day 1', 'Day 2', 'Day 3', 'Day 4'), each = length(day1))
df2 <- cbind(Subject, Data, Day)
```

When using `sm_forest()`, `x` is required to be a continuous variable. `y` should be a categorical variable (i.e., `factor`).

```{r,fig.width = 3.1, fig.height= 4.8, warning = FALSE}
ggplot(data = df2, aes(x = Value, y = Day, color = Day)) +
  sm_forest() 
```

`sm_forest()` has various `xxx.params` arguments as well as `...` generic parameter to control the aesthetics across various components of the plot.

- `avgPoint.params` sends arguments to `geom_point()` to plot the average point with specified `color`, `fill`, `size`, `shape` and `alpha`.

- `err.params = list()` sends arguments, such as `color`, to `geom_linerange()` to plot the error bar.

- `point.params = list()` sends arguments, such as `color`, `fill`, and `alpha` to `geom_point()` to plot the individual points.

- `refLine.params = list()` sends arguments, such as `color` and `size` to `geom_vline()` to plot the reference line.


We can reset all the defaults by listing these parameters as `xxx.params = list()`.

```{r, fig.width = 3.1, fig.height= 4.8, warning = FALSE}
ggplot(data = df2, aes(x = Value, y = Day, color = Day)) +
  sm_forest(avgPoint.params = list(),
            err.params = list(),
            point.params = list())
```

Let's restore the defaults and change colors now using `scale_color_manual()`.

```{r, fig.width = 3.1, fig.height= 4.8, warning = FALSE}
ggplot(data = df2, aes(x = Value, y = Day, color = Day)) +
  sm_forest() +
  scale_color_manual(values = sm_palette(4))
```


These points could be jittered (`point_jitter_width`) and increase the separation level between the average point and the individual points (`sep_level`), which has no limited range as opposed to 0-4 in `sm_raincloud()`.


```{r, fig.width = 3.1, fig.height= 4.8, warning = FALSE}
ggplot(data = df2, aes(x = Value, y = Day, color = Day)) +
  sm_forest(point_jitter_width = 0.12, sep_level = 3) +
  scale_color_manual(values = sm_palette(4))

```

If `sep_level` is positive, the average point is above the individual points. If `sep_level` is negative, then the average point is below the individual points.

```{r, fig.width = 3.1, fig.height= 4.8, warning = FALSE}
ggplot(data = df2, aes(x = Value, y = Day, color = Day)) +
  sm_forest(point_jitter_width = 0.12, sep_level = -3) +
  scale_color_manual(values = sm_palette(4))

```

The default error bar type is 95% confidence interval (`errorbar_type = 'ci'`). You can change it to either standard error (`'se'`) or standard deviation (`'sd'`). Here is standard deviation.

```{r, fig.width = 3.1, fig.height= 4.8, warning = FALSE}
ggplot(data = df2, aes(x = Value, y = Day, color = Day)) +
  sm_forest(point_jitter_width = 0.12, sep_level = 3,
            errorbar_type = 'sd') +
  scale_color_manual(values = sm_palette(4))

```

A common practice is to annotate the error range (95% confidence interval) in text on on the forest plot. This could be done using `sm_forest_annot()`.

```{r, fig.width = 3.1, fig.height= 4.8, warning = FALSE}
ggplot(data = df2, aes(x = Value, y = Day, color = Day)) +
  sm_forest(point_jitter_width = 0.12, sep_level = 3,
            errorbar_type = 'sd') +
  scale_color_manual(values = sm_palette(4)) +
  sm_forest_annot(data = df2, x = Value, y = Day, sep_level = 3)

```

`sm_forest_annot()` has various arguments. These arguments should be identical to what you have provided `ggplot(data = ..., aes(x = ..., y = ...)`.

- The `data` argument requires the variable that stores the data frame that is used to plot the data, which is `df2` in this example. 

- `x` argument is the variable that is plotted along the x-axis. 

- `y` argument is the variable that is plotted along the y-axis. 

Also, there are some additional arguments.

- `sep_level` refers to the separation level between the text annotation and the average point. If it is positive, the text will be above the mean point. **In practice, put the same value of** `sep_level` **in both** `sm_forest()` **and **`sm_forest_annot()`. The default is `sep_level = 2`, which is also found in `sm_forest()`.

- `errorbar_type` refers to the type of error by which the text annotation should be written. **In practice, put the same argument of** `sep_level` **in both** `sm_forest()` **and **`sm_forest_annot()`. The default is set to confidence interval (`errorbar_type = 'ci'`).

- `text.params` feeds arguments to `geom_text()` to set the aesthetics of the texts such as `color` and `size`. 

We can try `sep_level = -3` across `sm_forest()` and `sm_forest_annot()` to see what happens.

```{r, fig.width = 3.1, fig.height= 4.8, warning = FALSE}
ggplot(data = df2, aes(x = Value, y = Day, color = Day)) +
  sm_forest(point_jitter_width = 0.12, sep_level = -3,
            errorbar_type = 'sd') +
  scale_color_manual(values = sm_palette(4)) +
  sm_forest_annot(data = df2, x = Value, y = Day, sep_level = -3)

```

We see that the positions of the individual points and texts have been flipped in relation to the average points.

You can also draw a vertical reference line by setting `refLine = TRUE`.

```{r, fig.width = 3.1, fig.height= 4.8, warning = FALSE}
ggplot(data = df2, aes(x = Value, y = Day, color = Day)) +
  sm_forest(point_jitter_width = 0.12, sep_level = -3,
            errorbar_type = 'sd',
            refLine = TRUE) +
  scale_color_manual(values = sm_palette(4)) +
  sm_forest_annot(data = df2, x = Value, y = Day, sep_level = -3)

```

We could change the reference line's location using the argument `xintercept`.

```{r, fig.width = 3.1, fig.height= 4.8, warning = FALSE}
ggplot(data = df2, aes(x = Value, y = Day, color = Day)) +
  sm_forest(point_jitter_width = 0.12, sep_level = -3,
            errorbar_type = 'sd',
            refLine = TRUE,
            xintercept = 4) +
  scale_color_manual(values = sm_palette(4)) +
  sm_forest_annot(data = df2, x = Value, y = Day, sep_level = -3)

```

We could change the aesthetics of the reference line using `refLine.params`.

```{r, fig.width = 3.1, fig.height= 4.8, warning = FALSE}
ggplot(data = df2, aes(x = Value, y = Day, color = Day)) +
  sm_forest(point_jitter_width = 0.12, sep_level = -3,
            errorbar_type = 'sd',
            refLine = TRUE,
            xintercept = 4,
            ref.params = list(size = 1, alpha = 0.3, 
                              color = sm_color('darkred'))) +
  scale_color_manual(values = sm_palette(4)) +
  sm_forest_annot(data = df2, x = Value, y = Day, sep_level = -3)

```

### Plotting the forest plot using data of multiple groups

Plotting the forest plot using data of multiple groups is also possible with `sm_forest()`. To explore this feature, lets upload data from online.

```{r, warning = FALSE, message = FALSE}
df <- read_csv('https://www.smin95.com/amblyopia_random.csv')
df$SF <- as.factor(df$SF)
df1 <- filter(df, Condition == 'One') 
```

Now let's plot these data using `sm_forest()`. Make sure that group the data for each group `group = Group`.

```{r, fig.width = 3.1, fig.height= 4.8}
ggplot(df1, aes(x = absBP, y = SF, color = Group, group = Group)) +
  sm_forest(sep_level = 3) +
  scale_color_manual(values = sm_color('darkred', 'viridian')) +
  ggtitle('Forest plot of two groups') 
```

We could also set the transparency of the points so that the `alpha` increases over `SF`.

```{r, fig.width = 3.1, fig.height= 4.8}
ggplot(df1, aes(x = absBP, y = SF, color = Group, alpha = SF, group =Group)) +
  sm_forest(sep_level = 3) +
  scale_color_manual(values = sm_color('darkred', 'viridian')) +
  scale_alpha_manual(values = c(0.3,0.5,0.7,0.85, 1)) +
  ggtitle('Forest plot of two groups')
```

The issue with this plot is that the individual points are overlapped, as well as the mean points at `0.5` on the y-axis `SF`. You could use `position_dodge()` to `dodge` the position of one group to prevent the overlap.

```{r, fig.width = 3.1, fig.height= 4.8}
ggplot(df1, aes(x = absBP, y = SF, color = Group, alpha = SF, group =Group)) +
  sm_forest(sep_level = 3,
            position = position_dodge(width = .8)) +
  scale_color_manual(values = sm_color('darkred', 'viridian')) +
  scale_alpha_manual(values = c(0.3,0.5,0.7,0.85, 1)) +
  ggtitle('Forest plot of two groups')
```

As in `sm_raincloud()`, we could split the data frames into two and plot them separately. 

```{r}
df1amb <- filter(df1, Group == 'Amblyopia')
df1norm <- filter(df1, Group == 'Normal')
```

We could plot each group separately using the two data frames. First, plot the `Amblyopia` group.

```{r, fig.width = 3.1, fig.height= 4.8}

ggplot(mapping = aes(x = absBP, y = SF, color = Group, fill = Group)) +
  sm_forest(data = df1amb) +
  scale_color_manual(values = sm_color('darkred', 'viridian')) 
```

Next, apply `position_nudge()` to move all the components.

```{r, fig.width = 3.1, fig.height= 4.8}

ggplot(mapping = aes(x = absBP, y = SF, color = Group, fill = Group)) +
  sm_forest(data = df1amb, position = position_nudge(y = +0.25)) +
  scale_color_manual(values = sm_color('darkred', 'viridian')) 
```

We see that `sep_level = 2` no longer works when we apply `position_nudge()` here because `sep_level` also controls the position of the components using `position_nudge()`. We have bypassed its control by explicitly writing `position_nudge()`. Therefore, we will need to separate the points from the mean points manually by using `point.params`. 


```{r, fig.width = 3.1, fig.height= 4.8}

ggplot(mapping = aes(x = absBP, y = SF, color = Group, fill = Group)) +
  sm_forest(data = df1amb, position = position_nudge(y = +0.4),
            point.params = list(size = 2,
                                   show.legend = FALSE, alpha = 0.3,
                                   position = position_nudge(y = 0.2))) +
  scale_color_manual(values = sm_color('darkred', 'viridian')) 
```

Then, we can plot the data of `Normal` on top of the `Amblyopia` plot.

```{r, fig.width = 3.1, fig.height= 4.8}

ggplot(mapping = aes(x = absBP, y = SF, color = Group, fill = Group)) +
  sm_forest(data = df1amb, position = position_nudge(y = +0.3),
               point.params = list(size = 2, alpha = 0.3,
                                   position = position_nudge(y = 0.1))) +
  scale_color_manual(values = sm_color('darkred', 'viridian')) +
  sm_forest(data = df1norm,
               position = position_nudge(y = -0.3),
               point.params = list(size = 2,
                                   show.legend = FALSE, alpha = 0.3,
                                   position = position_nudge(y = -0.1)))  +
  ggtitle('Forest plot of two groups') 
```

This looks acceptable. Another solution is to just remove the individual points altogether by setting `sm_forest(points = FALSE)`. In this method, we do not need to split the data frame into two.

```{r, fig.width = 3.1, fig.height= 4.8}
ggplot(df1, aes(x = absBP, y = SF, color = Group, alpha = SF, group = Group)) +
  sm_forest(sep_level = 3,
            position = position_dodge(width = .8), points = FALSE) +
  scale_color_manual(values = sm_color('darkred', 'viridian')) +
  scale_alpha_manual(values = c(0.3,0.5,0.7,0.85, 1)) +
  ggtitle('Forest plot of two groups')
```

Also, `sm_forest_annot()` can be used to provide text annotations for the error range when presenting data of multiple groups. However, first, we will need to separate the two data frames as shown below for each group: `Amblyopia` and `Normal`.

```{r}
df1amb <- filter(df1, Group == 'Amblyopia')
df1norm <- filter(df1, Group == 'Normal')
```

Next, we will have to `position_nudge` for each annotation and group. There are five numbers that are provided in `position_nudge()`; the first one is for `SF = 0.5`, and the last one is for `SF = 8`. 

```{r, fig.width = 3.1, fig.height= 4.8}

ggplot(mapping = aes(x = absBP, y = SF, color = Group)) +
  sm_forest(data = df1amb, position = position_nudge(y = +0.2),
              points = FALSE) +
  sm_forest_annot(data = df1amb, x = absBP, y = SF,
                    position = position_nudge(y = c(0.2,-0.2,-0.2,-0.2,-0.2), 
                                            x = c(0.4,0,0,0,0)),
                  text.params = list(color = sm_color('darkred'), 
                                     size = 3)) +
  scale_color_manual(values = sm_color('darkred', 'viridian')) +
  sm_forest(data = df1norm,
               position = position_nudge(y = -0.2),
               points = FALSE)  +
  sm_forest_annot(data = df1norm, x = absBP, y = SF,
                  position = position_nudge(y = c(-0.18,0.2,0.2,0.2,0.2), 
                                            x = c(0.4, 0, 0, 0,0)), 
                  text.params = list(color = sm_color('viridian'),
                                     size = 3)) +
  ggtitle('Forest plot of two groups')
```

Here is the forest plot with data from multiple groups with text annotations but without individual points. As you may realize, text annotations can be extremely tedious so much so that I personally think using Adobe Illustrator might be better in this case. 