--- 
title: "Data Analysis and Visualization in R Using smplot"
author: "Seung Hyun Min"
date: "Last Updated: `r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
---

```{r include=FALSE,cache=FALSE,purl=FALSE}
# don't cache anything on this page
knitr::opts_chunk$set(cache = FALSE)
```

# Welcome! {-#index}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.4, fig.height= 11.7,warning = F}
library(tidyverse)
library(cowplot)
library(smplot)
df <- read_csv('https://www.smin95.com/data.csv') 
df$Day <- factor(df$Day, levels = c('One','Two','Three'))
ggplot(data = df, mapping = aes(x = Day, y = Value, fill = Day)) + 
  sm_boxplot(color = 'white', shape = 21) + 
  scale_fill_manual(values = sm_palette(3)) +
  ggtitle('Boxplot') + 
  theme(plot.title = element_text(face="bold")) -> boxplot
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_violin() + 
  scale_color_manual(values = sm_palette(3)) + 
  ggtitle('Violin plot') + 
  theme(plot.title = element_text(face="bold"))-> violin
set.seed(1) # generate random data
day1 = rnorm(20,0,1)
day2 = rnorm(20,5,1)
Subject <- rep(paste0('S',seq(1:20)), 2)
Data <- data.frame(Value = matrix(c(day1,day2),ncol=1))
Day <- rep(c('Day 1', 'Day 2'), each = length(day1))
df1 <- cbind(Subject, Data, Day)
ggplot(data = df1, mapping = aes(x = Day, y = Value, fill = Day)) +
  sm_bar(shape = 21, color = 'white', point_alpha = 1, bar_alpha = 0.2,
          bar_fill_color = 'black') +
  scale_fill_manual(values = sm_color('blue','orange')) +
  theme(axis.title.x = element_blank()) +
  ggtitle('Bar plot') + 
  theme(plot.title = element_text(face="bold")) -> bar
ggplot(data = df1, mapping = aes(x = Day, y = Value, 
                                group = Subject, fill = Day)) +
  sm_slope(labels = c('Day 1', 'Day 2'),
            shape = 21, color = 'white', line_alpha = 0.3) +
  scale_fill_manual(values = sm_color('blue','orange')) +
  ggtitle('Slope chart') + 
  theme(plot.title = element_text(face="bold")) -> slope
# raincloud plot
set.seed(2) # generate random data
day1 = rnorm(20,0,1)
day2 = rnorm(20,5,1)
day3 = rnorm(20,6,1.5)
day4 = rnorm(20,7,2)
Subject <- rep(paste0('S',seq(1:20)), 4)
Data <- data.frame(Value = matrix(c(day1,day2,day3,day4),ncol=1))
Day <- rep(c('Day 1', 'Day 2', 'Day 3', 'Day 4'), each = length(day1))
df2 <- cbind(Subject, Data, Day)
sm_raincloud(data = df2, x = Day, y = Value,
              boxplot_alpha = 0.5, color = 'white', shape = 21,
              sep_level = 2) +
  scale_x_continuous(limits = c(0.25,4.75), 
                     labels = c('1', '2', '3', '4'),
                     breaks = c(1,2,3,4)) +
  scale_color_manual(values = rep('transparent',4)) + 
  scale_fill_manual(values = sm_palette(4)) + 
  ggtitle('Raincloud plot') + 
  xlab('Day') + 
  theme(plot.title = element_text(face="bold")) -> raincloud
add_four <- plot_grid(boxplot, violin, bar, slope,
                 ncol =2,
                 nrow = 2, rel_widths = c(1,1,1,1),
                 scale = 0.95)
all <- plot_grid(add_four, raincloud, ncol = 1, 
                 nrow = 2, rel_widths = c(1,1), 
                 rel_heights = c(2,1.05))
print(all)
```

This book (https://smin95.com/dataviz) guides the reader to be familiar with R, a programming language, for data visualization. It also introduces the R package **smplot**, which aims to make the process of data visualization simple. This book was created entirely using RMarkdown for reproducibility. 

## Major updates on smplot and smCSF {-}

**smplot**: Latest updates have been made on **August 20th, 2022**. Please **re-download smplot**. These include： 

- Changes/fixes in `sm_minimal()` and `sm_raincloud()`.

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github('smin95/smplot', force = TRUE)
devtools::install_github('smin95/smCSF', force = TRUE)
```

## Please cite this paper when smplot is used {-}

**Seung Hyun Min, Jiawei Zhou. smplot: An R Package for Easy and Elegant Data Visualization (2021). Frontiers in Genetics, 12:802894.**

## For R and ggplot2 experts {-}

Please read Chapter 4, which is a manual for **smplot**.

## Issues and contact {-}
If you spot any mistakes in this online book, please pull-request on [Github](https://github.com/smin95/dataviz) or email me seung.min@mail.mcgill.ca. If you find issues with the 'smplot' package, please pull-request on [Github](https://github.com/smin95/smplot).

## License {-}
smplot is under the MIT license. This tutorial is licensed under a [Creative Commons Attribution 4.0 International License](https://creativecommons.org/licenses/by/4.0/). Seung Hyun Min is the sole creator of **smplot**.

## About the author {-}
I am Assistant Professor in Dept. Ophthalmology and Optometry at Wenzhou Medical University. On June 2021, I received my PhD at McGill University under the supervision of Dr. Robert F. Hess. My research is focused on amblyopia, neural plasticity and binocular vision. More details, such as my publications, are on my website: https://www.smin95.com. 
