---
output: html_document
chapter: Boxplot and Violin Plot
editor_options:
  chunk_output_type: console
---

# Boxplot and Violin Plot

Load these packages by typing the codes below.

```{r, warning = FALSE, message = FALSE}
library(tidyverse) # it has ggplot2 package
library(cowplot) # it allows you to save figures in .png file
library(smplot2)
```

## Upload sample data (csv file) in RStudio

### Download requirements 

Download **data.csv** from
https://www.smin95.com/data.csv

Put the file **data.csv** in your **LearnR** folder, where you will create a new script using RStudio.

### Before we get started 

First, within RStudio, set **LearnR** folder as your working directory by clicking: **Session -> Set Working Directory -> Choose Directory**. 

Then, load these three packages with the `library()` function: **ggplot2**, **cowplot** and **smplot2**.

The **tidyverse** package includes the **ggplot2** package. For this reason, instead of loading **ggplot2**, you can load **tidyverse** instead. **tidyverse** has other packages beside **ggplot2**, such as **dplyr**, which is also necessary to handle and manage data.

Remember from the last chapter that:

- Functions from the **ggplot2** package include `ggplot()` and other **geom** functions such as `geom_point()`, `geom_smooth()`, etc.

- Functions from the **smplot2** package improve the **ggplot2** graphs visually. However, they are not required to make a graph using **ggplot2**.

- The **cowplot** package has a function called `save_plot`, which is very useful for saving a graph as an image.
  
### data.csv 

To start off, let's load **data.csv** in R from your **LearnR** folder. Then store the loaded data into a variable called `df`. 

We will use a function from the **tidyverse** package called **read_csv()**.
  - For more information, type `?read_csv()`.

```{r}
df <- read_csv('data.csv')
```

If you get an **error** such as `object 'data.csv' not found`, make sure you set your working directory in the **LearnR** folder. Also make sure you have downloaded your data and saved it in the **LearnR** folder.

You can view the entire data frame by typing `View(df)`.

```{r, eval = F}
View(df)
```


There are **three columns** in **df**.

- The first column (**Subject**) is the subject number. S1 means the first subject, and S2 second.

- The second column (**Value**) is the data. It contains numbers.

- The third column (**Day**) is the day of testing. There are three days of measurement: **One**, **Two** and **Three**.


**What is the data about?**

- 30 patients were tested across three days to see if there was an improvement in their visual function (**Value**). The higher value, the better the vision.

- Day **One**: first day of the study. This is the baseline value. After this measurement, patients received treatment (ex. drug, surgery, etc).

- Day **Two**: second day of the study. This is the one day after receiving the treatment.

- Day **Three**: third day of the study. This is two days after receiving the treatment.

As some of you may have noticed, the data are all made from statistical simulations (i.e., fake data).

### `read_csv('data.csv')` 

Let's use the variable **df** to store the output from **data.csv**. 

**df** is shortened from **data frame**, which refers to a structure of the data that is commonly used for data analysis.

In fact, 'data.csv' has the structure of a typical **data frame**.

- Each **column** is a **variable**.

- Each **row** represents an **observation** 

In future, I suggest you save data that has a similar structure to that of **df**. 

- Put your own data in Microsoft Excel and save it as a **csv file** to load it on R.

**Exercise**: Open **data.csv** using Microsoft Excel. Change the name of third column into **Group**. Then, save it as **data2.csv**. What does this data frame indicate now?

It indicates there are 30 individuals for each group. Also, there are 3 groups, so there are 90 individuals total. So, it might be better to change the subject number so each person has a unique label, such as **S1** to **S90**. 
  
### Best way to plot data 

A popular way to visualize data is plotting a bar graph. However, a bar graph does not show the distribution of the data well.For this reason, boxplot and violin plots have become more popular in scientific writing. 

## Boxplot

### What is a boxplot? 

This is the basic structure of the boxplot.

```{r, echo = FALSE,fig.width = 5*6/4.2}
set.seed(5)
y <- c(rnorm(100), 4)
s <- boxplot.stats(y)

random_df <- data.frame(y = c(s$stats, max(y)),
                 x = c(1.23, 1.605, 1.605, 1.605, 1.23, 1.24),
                 label = c("minimum", "first quartile", "median", "third quartile", "maximum within upper whisker", "outlier"))

bp <- ggplot(data.frame(y), aes(x = 1, y = y)) + geom_boxplot(fill = sm_color('darkred')) +
  geom_text(data = random_df, aes(x, y, label = label), hjust = 0, size = 14/.pt) +
  scale_x_continuous(limits = c(0, 3.5), expand = c(0, 0)) +
  scale_y_continuous(limits = c(-2.55, 4.2), expand = c(0, 0)) +
  theme_nothing()

points <- ggplot(data.frame(y), aes(x = 0, y = y)) + 
  geom_point(color = sm_color('darkred'), position = position_jitter(width = .4, height = 0, seed = 320)) +
  scale_x_continuous(limits = c(-1.8, .4), expand = c(0, 0)) +
  scale_y_continuous(limits = c(-2.55, 4.2), expand = c(0, 0)) +
  theme_nothing()

plot_grid(points, bp, rel_widths = c(.65, 1), nrow = 1)
```

- The line in the middle is the **median** of the data.
- The **first quartile** is the lower 25% of the data.
- The **third quartile** is the 75% percentile of the data. 
- The **interquartile range** is the distance between the first and third quartiles.
- The **lower whisker** is the lowest data point without outliers.
- The **upper whisker** is the highest data point without outliers.
- The point that lies outside the whisker is the **outlier**.


### Boxplot vs bar graph 

Here is the basic structure of the bar graph.

```{r, echo = FALSE,fig.width = 5*6/4.2}
set.seed(5)
y <- c(rnorm(100,5,3))
avg <- mean(y)
se <- sd(y)/sqrt(length(y))

y_df <- data.frame(Avg = mean(y))

random_df <- data.frame(y = c(avg-se, avg, avg+se),
                 x = c(1.5, 0.1, 1.5),
                 label = c("lower standard error", "mean", "upper standard error"))

bar <- ggplot(y_df, aes(x = 1, y = Avg)) + geom_bar(stat = 'identity',fill = sm_color('darkred'), width = .8) +
  geom_text(data = random_df, aes(x, y, label = label), hjust = 0, size = 5) +
  scale_x_continuous(limits = c(0, 3.5), expand = c(0, 0)) +
  scale_y_continuous(limits = c(-2, 13), expand = c(0, 0)) +
  theme_nothing() +
  geom_errorbar(aes(ymin = Avg - se, ymax = Avg + se),
                width = .1, size = 1.2)

points <- ggplot(data.frame(y), aes(x = -0.2, y = y)) + 
  geom_point(color = sm_color('darkred'), position = position_jitter(width = .4, height = 0, seed = 320)) +
  scale_x_continuous(limits = c(-1.8, .4), expand = c(0, 0)) +
  scale_y_continuous(limits = c(-2, 13), expand = c(0, 0)) +
  theme_nothing()

plot_grid(points, bar, rel_widths = c(.65, 1), nrow = 1)
```

- The boxplot is more informative than bar graph.
- The bar graph only shows the mean (average) of the data and the standard error.
- The bar graph does not represent the distribution very well. The narrow errorbar is quite misleading here.
- For this reason, let's focus on drawing a boxplot.


### Boxplot using ggplot2 

We want to plot **Value** (y-axis) from each **Day** (x-axis). As always in **ggplot**, let's start by writing `ggplot()`.

Make sure you specify the data to plot by writing `data = df` in `ggplot()`. Also, you will need to specify the x- and y-axes in the `mapping` argument within `ggplot()`.

- `x = Day` and `y = Value`.
  
```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value))
```

However, as we learned in the last lesson, **ggplot** requires **geom** objects to plot something. Last time we used **geom_point()**, which plots points.
- Now, let's use **geom_boxplot()** to plot a boxplot.

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value)) + geom_boxplot()
```

Here's the boxplot! But I think there is a problem here. Day **Three** precedes Day **Two** in the x-axis. 

- **Exercise**: Why do you think **Three** appeared first?

### Problem: Alphabetical order 

The issue with the boxplot is that the x-axis follows an alphabetical order.

- **Th** precedes **Tw**. In short, R does not seem to know that we want the order to be **One -> Two -> Three**.

R has decided to follow the alphabetical order is because the column **Day** is shown as `<chr>`, which means **character**.

```{r}
df
```

- Note that the **Subject** column is also considered as `<chr>` (**characters**) because it contains texts/alphabets.
- The **Value** column is categorized as `<dbl>`, which means **double**. 
  - **Double** basically means numbers.

In R, **character** is also referred to as **string**. They mean the same thing. Here are more examples of **characters/strings**.

```{r}
a <- 'R is very useful!'
b <- 'Dogs are so cute.'
c <- 'Clarinet is the best.'
```

In these examples, you have stored the **characters/strings** in the variables **a**, **b** and **c**.

### Displaying characters in a non-alphabetical order 

**Factors** are important when you want to display characters in non-alphabetical order. **Factors** are variables that have a limited number of different values. 

For example, if a dataset has two groups (**controls** and **patients**), these can be referred to as **factors**.

However, when you are using `read_csv()` to load data into RStudio, `read_csv()` will interpret texts/alphabets as **characters**, not **factors**.

- This is a problem when you need a non-alphabetical order, such as **one -> two -> three** and  **Jan -> Feb -> Mar -> Apr**. 

To resolve this problem, you will need to change the **characters** into **factors** using `factor()` function.

```{r}
df$Day <- factor(df$Day, levels = c('One','Two','Three'))
```

Remember that `<-` means equivalent. Also,  `$` of a data frame can extract a specific column 

- You can also type `df$Value` and run it to see what happens. You will get all values of the second column **Value**.

```{r, eval = F}
df$Value
```

By using `factor()`, you are making the **Day** column of **df** into factor from character. However, you will need to store the **factored** version of `df$Day` by using `<-`. If not, `df$Day` will stay as **character**.

`levels` argument in `factor()` allows you to specify the order of the factors.

For more information about **factors**, please check Chapter 15 of *R for Data Science* (https://r4ds.had.co.nz/factors.html).

- **Note:** you do not have to understand this part 100%. However, if you do encounter similar issues, I suggest you resolve it with the solution above.

### Double-check if the 'Day' column is factor 

- Let's open the data frame **df** again.

```{r}
df
```

Now we see `<fct>` in the **Day** column. This means that the column is now **factor**, not **character**.

### Back to boxplot using ggplot2 

Now let's draw the boxplot again.

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value)) + geom_boxplot()
```

Since we have converted the **Day** column of **df** into factor, the boxplot successfully plots each day in the correct order.

### A boxplot with individual data 

Sometimes, it is preferable to plot individual data on top of the boxplot. There are many ways to do this. Here is a simple solution by using `geom_jitter()`. For more information, type `?geom_jitter`.

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value)) + geom_boxplot() + 
  geom_jitter()
```

I think the points are too spread. So let's set the `width` of the spread in `geom_jitter()` to 0.15. Also, let's apply a different color of points for each **Day**.

```{r,fig.width = 4.8, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  geom_boxplot() + 
  geom_jitter(width = 0.15)
```

Now we can make the boxplot prettier with `sm_minimal()`, which removes all grids in the graph. We can also set `legends = FALSE` to remove the legends.

Let's also add a title **Visual improvement after treatment** using `ggtitle()`, which is a function from the **ggplot2** package.

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  geom_boxplot() + 
  geom_jitter(width = 0.15) + 
  sm_minimal(legends = FALSE) +
  ggtitle('Visual improvement after treatment')
```

Now let's set the colors using `sm_palette()`. Remember from the last lesson that we set the colors in **ggplot** by using `scale_color_manual()` for points without borders.

- Let's write `3` in `sm_palette()` because we need 3 colors (for each **Day**). It will automatically extract 3 different colors from the color palette in **smplot2**. 

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  geom_boxplot() + 
  geom_jitter(width = 0.15) + 
  sm_minimal(legends = FALSE) +
  ggtitle('Visual improvement after treatment') +
  scale_color_manual(values = sm_palette(3))
```

If you need horizontal grids, you can simply replace `sm_minimal()` with `sm_hgrid()`. There is also `sm_vgrid()` but I will not try it here because it is not so useful in our context.


```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  geom_boxplot() + 
  geom_jitter(width = 0.15) + 
  sm_hgrid(legends = FALSE) +
  ggtitle('Visual improvement after treatment') +
  scale_color_manual(values = sm_palette(3))
```

### `geom_boxplot()` vs `sm_boxplot()` 

`geom_boxplot()` does not show individual data. You need to add `geom_jitter()`.

`sm_boxplot()` shows both a boxplot and jittered individual data. It automatically uses `sm_hgrid()`. Also it stores defaults that are aesthetically pleasing and has flexibility with which you can entirely overturn the defaults.

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_boxplot() + 
  ggtitle('Visual improvement after treatment') +
  scale_color_manual(values = sm_palette(3))
```

As you can see the default of the point is slightly transparent (`alpha = 0.65'`). If you prefer it to change it to opaque with white borders, we can by specifying the parameters of the points, `point.param`, which is an argument that is found across multiple functions of **smplot2**.

In arguments that look like`xxx.params = list()`. `sm_boxplot()` has two `xxx.param`s: `point.param` and `boxplot.param`. 

- Parameters in `point.param` are used in `geom_point()` to plot the points.

- Parameters in `boxplot.param` are used in `geom_boxplot()` to plot the boxplot.

In the default of `sm_boxplot()`, the shape of the point is `16`. This information can be accessed by seeing the defaults within the `help` page of `?sm_boxplot`. 

We could change the shape to `21` so that the points have the circular shape with borders. In the case of `shape = 21`, `fill` decides the filling color rather than `color`, which sets the color of the border. Therefore, the variable in `aes()` has to be changed so that there is unique filling color for each day as `fill = Day`.

Also, `scale_color_manual()` is now used to apply distinct border colors of the points rather than their filling colors. So now we should use `scale_fill_manual()`.

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, fill = Day)) + 
  sm_boxplot(point.params = list(shape = 21, color = 'white',
                                 size = 2.5)) + 
  ggtitle('Visual improvement after treatment') +
  scale_fill_manual(values = sm_palette(3))
```

Notice that when you leave `point.params = list()` as empty or filled, then the aesthetic defaults of `sm_boxplot()` will disappear, such as the transparency of the points (`alpha = 0.65`). To fully restore the defaults while keeping the shape of the points, you will need to specify all the parameters.

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, fill = Day)) + 
  sm_boxplot(point.params = list(shape = 21, color = 'white',
                                 size = 2.5, alpha = 0.65)) + 
  ggtitle('Visual improvement after treatment') +
  scale_fill_manual(values = sm_palette(3))
```

Also, this is an example that shows that the defaults (i.e., transparency) are no longer maintained when `point.params = list()` is written but not filled. 

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_boxplot(point.params = list()) + 
  ggtitle('Visual improvement after treatment') +
  scale_color_manual(values = sm_palette(3))
```

Another way to change the shape is to not use `point.params = list()`. You could directly use `shape` argument within `sm_boxplot()` but this could affect other aesthetic components.

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_boxplot(shape = 15) + 
  ggtitle('Visual improvement after treatment') +
  scale_color_manual(values = sm_palette(3))
```

In this case, we are lucky because `shape` is an aesthetic parameter that does not affect the boxplot, which is drawn using `geom_boxplot()`. We can set the size and transparency explicity in `point.params()`. 


```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_boxplot(shape = 15, point.params = list(alpha = 0.65, 
                                             size = 2.5)) + 
  ggtitle('Visual improvement after treatment') +
  scale_color_manual(values = sm_palette(3))
```

What happens if you explicittly write `boxplot.params = list()` in `sm_boxplot()`? 

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_boxplot(boxplot.params = list()) + 
  ggtitle('Visual improvement after treatment') +
  scale_color_manual(values = sm_palette(3))
```

Now the defaults of the boxplot are gone. The boxplot's `fill` is white, and the line colors are no longer `black`. Also, the `width` is wider now than the default (`width = 0.5`).

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_boxplot(boxplot.params = list(fill = 'gray95', width = 0.5, 
                                   color = 'black')) + 
  ggtitle('Visual improvement after treatment') +
  scale_color_manual(values = sm_palette(3))
```

You can remove the label of the outlier by specifying `outlier.shape = NA`, which is an argument within `geom_boxplot()`, in `boxplot.params`.

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_boxplot(boxplot.params = list(fill = 'gray95', width = 0.5, 
                                   color = 'black', outlier.shape = NA)) + 
  ggtitle('Visual improvement after treatment') +
  scale_color_manual(values = sm_palette(3))
```

Now we are essentially back to where we were.

In addition, you can also change the `size` as a generic aesthetic parameter without affecting others because the `size` of the boxplot (`boxplot.params`) already has a specific default. 


```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_boxplot(shape = 15, size = 2) + 
  ggtitle('Visual improvement after treatment') +
  scale_color_manual(values = sm_palette(3))
```

**Important:** The generic parameter (`...` in ex. `sm_boxplot()`) cannot overturn the smplot2's default unless the user makes the default gone by explictly writing `param.xxx = list()`.

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_boxplot(shape = 15, size = 2,
             boxplot.params = list()) + 
  ggtitle('Visual improvement after treatment') +
  scale_color_manual(values = sm_palette(3))
```

Now the `size` affects both the points and the boxplot because the default of `boxplot.params` is gone after we have written it in the argument.

Advanced users can really have much flexibility with this feature of **smplot2**. For example, you can essentially remove all the defaults by including empty `boxplot.params` and `point.params`.

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_boxplot(boxplot.params = list(),
             point.params = list()) + 
  ggtitle('Visual improvement after treatment') +
  scale_color_manual(values = sm_palette(3))
```

Then you can change the `color` across points and boxplot to be just `black`.

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_boxplot(color = 'black', boxplot.params = list(),
             point.params = list()) + 
  ggtitle('Visual improvement after treatment') +
  scale_color_manual(values = sm_palette(3))
```

If you are confused about this, do not worry because you can directly set the `size` and `shape` of the points as generic parameter without affecting other aesthetic components. You could probably get by with the defaults of **smplot2** without specifying much. Essentially, this was up **smplot** was able to offer most.

Now, pick your favorite boxplot and store it in a variable called **my_boxplot** using `<-`.

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
my_boxplot <- ggplot(data = df, mapping = aes(x = Day, y = Value, fill = Day)) + 
  sm_boxplot(point.params = list(size = 2.5, color='white', shape = 21)) + 
  ggtitle('Visual improvement after treatment') +
  scale_fill_manual(values = sm_palette(3))
```

### Save boxplot 

We can save the boxplot by using `save_plot()` from the **cowplot** package.

Let's set the aspect ratio (width/height) to 1. So, type `base_asp = 1`.

```{r, eval = F}
save_plot('boxplot.jpg', my_boxplot, base_asp = 1)
```

This image file should now be in your **LearnR** folder.


## Violin Plot 

The boxplot represents the distribution quite well but it can still be misleading.Also, it can be visually overwhelming (i.e., too crowded) to some readers. For this reason, researchers have begun favoring the violin plot more.

### Structure of the violin plot 

```{r, fig.width = 5*6/4.2, echo = F, warning=F}
set.seed(7)
y <- c(rnorm(100), 3.4)
d <- density(y)
random_df <- data.frame(y = c(min(y), d$x[which(d$y == max(d$y))], max(y)), x = c(1.3, 1.7, 1.3),
                 label = c("minimum data", "maximum point density", "maximum data"))
p_violin <- ggplot(data.frame(y), aes(x = 1, y = y)) + geom_violin(fill = sm_color('darkred'), color = sm_color('darkred')) +
  geom_text(data = random_df, aes(x, y, label = label), hjust = 0,
            size = 14/.pt) +
  scale_x_continuous(limits = c(0, 3.5), expand = c(0, 0)) +
  scale_y_continuous(limits = c(-2.55, 3.6), expand = c(0, 0)) +
  theme_nothing()
p_points <- ggplot(data.frame(y), aes(x = 0, y = y)) + 
  geom_point(color = sm_color('darkred'), position = position_jitter(width = .4, height = 0, seed = 320)) +
  scale_x_continuous(limits = c(-1.8, .4), expand = c(0, 0)) +
  scale_y_continuous(limits = c(-2.55, 3.6), expand = c(0, 0)) +
  theme_nothing()
plot_grid(p_points, p_violin, rel_widths = c(.65, 1), nrow = 1)
```

- The part in the middle is the widest because it has most points. This part has the highest **point density**.
- The lower and upper ends of the violin plot are thinner because there are less points.
- Let's draw a violin plot in R.

### Violin plot with ggplot2 

Let's draw the violin plot. Specify the x-axis to **Day** and the y-axis to **Value** using data frame **df**.

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value)) + geom_violin()
```

Notice that since we have already converted the **Day** column of df into **factor**, the violin plot's x-axis is in the correct order. Also, notice that we can just replace `geom_boxplot()` with `geom_violin()`.

### Violin plot with individual points 

Let's add individual points to the violin plot by using `geom_jitter()`. 
- Set `width` of the spread to `0.15`.

```{r,fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value)) + 
  geom_violin() + 
  geom_jitter(width = 0.15)
```

Now let's apply different color to each **Day** like we did in boxplot by using `scale_color_manual()`.
- Let's use `sm_palette(3)`.
  
```{r,fig.width = 4.54, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  geom_violin() + 
  geom_jitter(width = 0.15) +
  scale_color_manual(values = sm_palette(3))
```

Finally, let's make it pretty and remove the **legend** by applying `sm_hgrid()` theme, which only has horizontal grids.
- Let's also add a title **Visual improvement after treatment** using `ggtitle()`.

```{r, fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) +
  geom_violin() + 
  geom_jitter(width = 0.15) +
  scale_color_manual(values = sm_palette(3)) +
  sm_hgrid() +
  ggtitle('Visual improvement after treatment')
```

### geom_violin() vs sm_violin() 

- `geom_violin()` does not show individual data. You need to add geom_jitter().
- `sm_violin()` shows both a violin plot and individual data.
  - It automatically uses `sm_hgrid()`.
  - It also display lines that represent the mean and +/- standard deviation. 

In `sm_violin()`, the default shape (`'point_shape'`) is 16, which is the circle without borders. Therefore, `color` fills the color of the individual points rather than `fill`. Also, `color` affects the color of the lines that show the mean and +/- standard deviation as well as the average point. 

Let’s set `color = Day` in `aes()` so that we can apply different colors of the lines and the points for each group.
- Make sure you use `scale_color_manual()` rather than `scale_fill_manual` to have a unique filling color for each `Day`.

```{r, fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_violin() + 
  scale_color_manual(values = sm_palette(3)) +
  ggtitle('Visual improvement after treatment')
```

As in the case of `sm_boxplot()`, you can specify the aesthetics of `sm_violin()` using `point.params`, `err.params` and `violin.params`. 

- `point.params` feeds the arguments to `geom_point()`, such as `color`, `alpha`, etc, to plot the individual points.

- `err.params` feeds the arguments to `geom_linerange()`, such as `color`, `size` etc, to plot the error bar in either standard error (`errorbar_type = se`), standard deviation (`errorbar_type = sd`) or 95% confidence interval (`errorbar_type = ci`).

- `violin.params` feeds the arguments to `geom_violin()`, such as `color` and `fill`, to plot the violin.

By using `size` and `shape` as generic parameter, you can easily change the mean and individual points' color and shapes, but not those of the violin because `geom_violin()` does not use these two parameters.

```{r, fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_violin(size = 3, shape = 15) + 
  scale_color_manual(values = sm_palette(3)) +
  ggtitle('Visual improvement after treatment')
```

In this case, we have successfully changed the size (larger) and shape (circle to square) of the mean and individual points. 

If you write `xxx.params = list()`, even with empty ones, you will remove the defaults of `sm_violin()`.

```{r, fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_violin(point.params = list(), 
            violin.params = list()) +
  ggtitle('Visual improvement after treatment')
```

Now we can add `color = 'black'` and `fill = 'black'` to affect it across all components of the graph (violin, error bar and points).

```{r, fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_violin(color = 'black', fill = 'black', 
            point.params = list(), 
            violin.params = list()) + 
  ggtitle('Visual improvement after treatment')
```

Only after the defaults are gone, the generic parameters (`...`) in `sm_violin()` have power to affect different components of the violin plot. However, if we add parameters within `xxx.params`, we see that these parameters will ignore the generic ones.

```{r, fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_violin(color = 'black', fill = 'black', 
            point.params = list(fill = 'white', shape = 21, 
                                color ='black', size = 2), 
            violin.params = list(alpha = 0.5)) + 
  ggtitle('Visual improvement after treatment')
```

You could also remove `borders` or add `legends` by setting them as `FALSE` or `TRUE` respectively (ex. `sm_violin(..., borders = FALSE)`).

```{r, fig.width = 3.5, fig.height=3.5, warning = F}
ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_violin(size = 2, shape = 17, borders = 'FALSE') + 
  scale_color_manual(values = sm_palette(3)) +
  ggtitle('Visual improvement after treatment')
```


### Save your violin plot 

Pick your favorite violin plot and store it in a variable called `my_violin` using `<-`.

```{r, evak = F}
my_violin <- ggplot(data = df, mapping = aes(x = Day, y = Value, color = Day)) + 
  sm_violin() + 
  scale_color_manual(values = sm_palette(3)) +
  ggtitle('Visual improvement after treatment')
```

We can save the violin plot by using `save_plot()` from the **cowplot** package, and set the aspect ratio (width/height) to 1. So, type `base_asp = 1`.

```{r, eval = F}
save_plot('violin.jpg', my_violin, base_asp = 1)
```

This image file should now be in your **LearnR** folder.

**Exercise**: `xxx.params` feeds arguments to `geom_xxx` functions. Understand the difference between the parameters that are within `xxx.params = list()` and `...`, which can be generic parameters. 

## Summary 

- You have learned how to convert **characters** to **factors**.
- **Factor** is important to order words/letters in a non-alphabetical order.
- You can now make a **boxplot** and a **violin plot** with R.
  - They are more informative than a bar graph.
- You have learned to use **geom** functions, such `geom_boxplot()`, `geom_violin()`, and `geom_jitter()`.
  - These are functions from the **ggplot2** package.
- You can also plot individual data at the same time with functions from the **smplot2** package.
  - These functions are `sm_boxplot()` and `sm_violin()`.
  